---
title: "Analysis"
author: "Nina Cutner, Kelsey Hammond, Sofia Favela"
format: pdf
execute:
  warning: false
  message: false
---

# Set Up

## Packages
```{r}
#| label: packages

# data wrangling and visualization
library(tidyverse)

# allows file paths to work across computers
library(here)

# formatting axis labels
library(scales)

# working with dates
library(lubridate)

# combining plots
library(patchwork)
```

## Data

```{r}
#| label: data

# read in bird survey data
birds <- read.csv(here("data", "birds.csv"))

# read in NOAA weather data
weather <- read.csv(here("data", "NOAA_daily_summaries.csv"))
```

# Preliminary Code for Analyses

## Cleaning Data

```{r}
#| label: cleaning-birds

birds_clean <- birds |>
  # convert observation date to date format
  mutate(
    observation_date = ymd(observation_date),

    # create monthly grouping variable
    month = floor_date(observation_date,
                       unit = "month")
  ) |>

  # filter to only include waterfowl observations
  filter(
    e_bird_group == "Waterfowl",

    # remove repeat observations
    repeat_observation == "No",

    # keep observations within study period
    observation_date > ymd("2019-09-30"),
    observation_date < ymd("2025-10-01")
  )
```

```{r}
#| label: cleaning-weather

weather_clean <- weather |> 
  
  # convert date column to date format
  mutate(
    DATE = ymd(DATE),
    
    # create monthly grouping variable
    month = floor_date(DATE,
                       unit = "month")
  ) |> 
  
  # filter to study period
  filter(
    DATE > ymd("2019-09-30"),
    DATE < ymd("2025-10-01")
  )
```

## Summaries

```{r}
#| label: monthly-precipitation

# calculate total monthly precipitation
monthly_precip <- weather_clean |> 
  
  # group observations by month
  group_by(month) |> 
  
  # sum precipitation within each month
  summarize(
    monthly_prcp = sum(PRCP, na.rm = TRUE),
    
    # remove grouping after summarizing
    .groups = "drop"
  )
```

```{r}
#| label: monthly-waterfowl-abundance

# calculate monthly abundance by species
monthly_waterfowl <- birds_clean |>
  group_by(month, species) |>
  summarize(
    abundance = sum(count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(species = factor(species, levels = c(
    "Snow Goose",
    "Ross's Goose",
    "Cackling Goose (Aleutian)",
    "Canada Goose",
    "Greater White-fronted Goose",
    "Ring-necked Duck",
    "Redhead",
    "Canvasback",
    "Ruddy Duck",
    "American Wigeon",
    "Blue-winged Teal",
    "Cinnamon Teal",
    "Northern Pintail",
    "Mallard",
    "Northern Shoveler",
    "Bufflehead",
    "Hooded Merganser",
    "Green-winged Teal",
    "Gadwall"
  )))
```

```{r}
#| label: total-monthly-waterfowl

# calculate total monthly waterfowl abundance
total_monthly_waterfowl <- birds_clean |>
  
  # group by month
  group_by(month) |>
  
  # sum all bird counts within each month
  summarize(
    total_abundance = sum(count, na.rm = TRUE),
    .groups = "drop"
  )
```

## Joining Data

```{r}
#| label: joining-monthly-data

# join monthly bird abundance with monthly precipitation
bird_weather_monthly <- total_monthly_waterfowl |>

  # join by month
  left_join(monthly_precip,
            by = "month") |>

  # arrange chronologically
  arrange(month) |>

  # create lagged precipitation and month-of-year variables
  mutate(

    # previous month's precipitation (for figure 3)
    lagged_prcp = lag(monthly_prcp),

    # numeric month of year 1-12 (for figure 3 color scale)
    month_of_year = month(month),

    # season label derived from calendar month
    # (Dec/Jan/Feb = Winter, Mar/Apr/May = Spring,
    #  Jun/Jul/Aug = Summer, Sep/Oct/Nov = Fall)
    season = case_when(
      month(month) %in% c(12, 1, 2) ~ "Winter",
      month(month) %in% c(3, 4, 5)  ~ "Spring",
      month(month) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                           ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))
  )
```

```{r}
#| label: seasonal-summary

# step 1: sum abundance and precipitation within each season-year combination
seasonal_by_year <- bird_weather_monthly |>

  # extract water year (Oct = start of new water year)
  mutate(
    water_year = if_else(month(month) >= 10,
                         year(month) + 1,
                         year(month))
  ) |>

  group_by(season, water_year) |>
  summarize(
    # total waterfowl abundance for this season in this year
    seasonal_abundance = sum(total_abundance, na.rm = TRUE),

    # total precipitation for this season in this year
    seasonal_prcp = sum(monthly_prcp, na.rm = TRUE),

    .groups = "drop"
  )

# step 2: average each season's totals across all water years
seasonal_summary <- seasonal_by_year |>

  group_by(season) |>
  summarize(
    avg_abundance = mean(seasonal_abundance, na.rm = TRUE),
    avg_prcp      = mean(seasonal_prcp, na.rm = TRUE),
    .groups = "drop"
  )
```

## Statistical Analysis Code

```{r}
#| label: linear-model

# linear regression: does previous month's precipitation
# predict total monthly waterfowl abundance?
abundance_model <- lm(total_abundance ~ lagged_prcp,
                      data = bird_weather_monthly)

summary(abundance_model)
```

# Visualizations

## Figure 1: Monthly Precipitation Through Time

```{r}
#| label: figure-1-precipitation
#| fig-width: 10
#| fig-height: 5

ggplot(data = monthly_precip,
       mapping = aes(x = month,
                     y = monthly_prcp)) +

  # shaded area beneath line adds visual depth; line shows trend
  geom_area(fill = "#4682B4", alpha = 0.25) +
  geom_line(color = "#7EC8E3", linewidth = 0.8) +

  # label x-axis by year only
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    expand = c(0, 0)
  ) +

  scale_y_continuous(expand = c(0, 0)) +

  labs(
    title = "Monthly Precipitation Through Time",
    x     = "Date",
    y     = "Monthly Precipitation (in)"
  ) +

  # custom dark marsh-inspired theme applied consistently across all figures
  theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = "#1C2B2D", color = NA),
    panel.background = element_rect(fill = "#243336", color = NA),
    panel.grid.major = element_line(color = "#3A4F52", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    plot.title  = element_text(color = "#F0EAD6", size = 20, face = "bold",
                               margin = margin(b = 12), hjust = 0),
    axis.text   = element_text(color = "#B8C9C2", size = 10),
    axis.title  = element_text(color = "#D4C9A8", size = 11),
    axis.ticks  = element_line(color = "#3A4F52"),
    plot.margin = margin(16, 20, 12, 16)
  )
```

## Figure 2: Monthly Waterfowl Abundance Through Time by Species

```{r}
#| label: figure-2-waterfowl
#| fig-width: 11
#| fig-height: 7

# manually assigned color palette for all 19 waterfowl species
waterfowl_colors <- c(
  "American Wigeon"             = "#C1440E",
  "Blue-winged Teal"            = "#1B6CA8",
  "Bufflehead"                  = "#E8A838",
  "Cackling Goose (Aleutian)"   = "#9DBF9E",
  "Canada Goose"                = "#6B8E23",
  "Canvasback"                  = "#B22222",
  "Cinnamon Teal"               = "#A0522D",
  "Gadwall"                     = "#8FBC8F",
  "Greater White-fronted Goose" = "#D2B48C",
  "Green-winged Teal"           = "#2E8B57",
  "Hooded Merganser"            = "#20B2AA",
  "Mallard"                     = "#4682B4",
  "Northern Pintail"            = "#5F9EA0",
  "Northern Shoveler"           = "#3CB371",
  "Redhead"                     = "#8B1A1A",
  "Ring-necked Duck"            = "#6A5ACD",
  "Ross's Goose"                = "#F4E4B8",
  "Ruddy Duck"                  = "#CD853F",
  "Snow Goose"                  = "#B0C4DE"
)

ggplot(data = monthly_waterfowl,
       mapping = aes(x = month,
                     y = abundance,
                     fill = species)) +

  # geom_col used instead of geom_area to avoid interpolation between
  # sparse survey dates (geom_area would create misleading filled regions
  # between months with no observations)
  geom_col(color = "#1C2B2D", linewidth = 0.1) +

  scale_fill_manual(values = waterfowl_colors) +

  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    expand = c(0, 0)
  ) +

  scale_y_continuous(expand = c(0, 0)) +

  labs(
    title = "Monthly Waterfowl Abundance Through Time",
    x     = "Date",
    y     = "Monthly Abundance (individuals)",
    fill  = NULL
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = "#1C2B2D", color = NA),
    panel.background = element_rect(fill = "#243336", color = NA),
    panel.grid.major = element_line(color = "#3A4F52", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    plot.title   = element_text(color = "#F0EAD6", size = 20, face = "bold",
                                margin = margin(b = 12), hjust = 0),
    plot.caption = element_text(color = "#7A9A95", size = 8, hjust = 0,
                                margin = margin(t = 8)),
    axis.text    = element_text(color = "#B8C9C2", size = 10),
    axis.title   = element_text(color = "#D4C9A8", size = 11),
    axis.ticks   = element_line(color = "#3A4F52"),
    legend.position   = "right",
    legend.background = element_rect(fill = "#1C2B2D", color = NA),
    legend.key        = element_rect(fill = NA, color = NA),
    legend.text       = element_text(color = "#F0EAD6", size = 8.5),
    legend.key.size   = unit(0.45, "cm"),
    legend.spacing.y  = unit(0.1, "cm"),
    plot.margin = margin(16, 20, 12, 16)
  )
```

## Figure 3: Lagged Precipitation vs. Total Waterfowl Abundance

```{r}
#| label: figure-3-lagged-precip
#| fig-width: 10
#| fig-height: 6

ggplot(data = bird_weather_monthly,
       mapping = aes(x = lagged_prcp,
                     y = total_abundance,
                     # color points by month of year to reveal seasonal structure
                     color = month_of_year)) +

  geom_point(alpha = 0.85, size = 4) +

  # turbo palette provides high contrast across the 1-12 month range
  scale_color_viridis_c(
    name   = "Month",
    option = "turbo",
    breaks = c(1, 3, 5, 7, 9, 11),
    labels = c("Jan", "Mar", "May", "Jul", "Sep", "Nov"),
    limits = c(1, 12)
  ) +

  labs(
    title   = "Relationship Between Precipitation and Waterfowl Abundance",
    x       = "Previous Month's Precipitation (in)",
    y       = "Total Monthly Waterfowl Abundance (individuals)",
    caption = "Each point represents one month's waterfowl abundance plotted
against the previous month's precipitation. Points colored by month of year."
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = "#1C2B2D", color = NA),
    panel.background = element_rect(fill = "#243336", color = NA),
    panel.grid.major = element_line(color = "#3A4F52", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    plot.title   = element_text(color = "#F0EAD6", size = 20, face = "bold",
                                margin = margin(b = 12), hjust = 0),
    plot.caption = element_text(color = "#7A9A95", size = 8.5, hjust = 0,
                                margin = margin(t = 8)),
    axis.text    = element_text(color = "#B8C9C2", size = 10),
    axis.title   = element_text(color = "#D4C9A8", size = 11),
    axis.ticks   = element_line(color = "#3A4F52"),
    legend.background = element_rect(fill = "#1C2B2D", color = NA),
    legend.key        = element_rect(fill = NA, color = NA),
    legend.text       = element_text(color = "#F0EAD6", size = 9),
    legend.title      = element_text(color = "#D4C9A8", size = 10, face = "bold"),
    legend.key.height = unit(1.2, "cm"),
    plot.margin = margin(16, 20, 12, 16)
  )
```

## Figure 4: Average Seasonal Trends Comparing Waterfowl Abundance and Precipitation

```{r}
#| label: figure-4-seasonal-abundance-precip
#| fig-width: 9
#| fig-height: 8

# shared theme object to keep both panels visually consistent
seasonal_theme <- theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = "#1C2B2D", color = NA),
    panel.background = element_rect(fill = "#243336", color = NA),
    panel.grid.major = element_line(color = "#3A4F52", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    plot.title  = element_text(color = "#F0EAD6", size = 14, face = "bold",
                               margin = margin(b = 8), hjust = 0),
    axis.text   = element_text(color = "#B8C9C2", size = 10),
    axis.title  = element_text(color = "#D4C9A8", size = 11),
    axis.ticks  = element_line(color = "#3A4F52"),
    plot.margin = margin(12, 20, 8, 16)
  )

# panel A: waterfowl abundance by season
# small jittered points = individual water-year totals; larger points + line = cross-year average
plot_abundance <- ggplot() +
  geom_point(
    data     = seasonal_by_year,
    mapping  = aes(x = season, y = seasonal_abundance, group = 1),
    color    = "#7EC8E3", alpha = 0.35, size = 2.5,
    position = position_jitter(width = 0.08, seed = 42)
  ) +
  geom_line(
    data    = seasonal_summary,
    mapping = aes(x = season, y = avg_abundance, group = 1),
    color   = "#7EC8E3", linewidth = 1.2
  ) +
  geom_point(
    data    = seasonal_summary,
    mapping = aes(x = season, y = avg_abundance, group = 1),
    color   = "#7EC8E3", size = 4
  ) +
  labs(
    title = "Average Total Seasonal Waterfowl Abundance",
    x     = "Season",
    y     = "Abundance (individuals)"
  ) +
  seasonal_theme

# panel B: precipitation by season
# same structure as panel A; olive green used to distinguish precipitation from bird data
plot_precip <- ggplot() +
  geom_point(
    data     = seasonal_by_year,
    mapping  = aes(x = season, y = seasonal_prcp, group = 1),
    color    = "#6B8E23", alpha = 0.35, size = 2.5,
    position = position_jitter(width = 0.08, seed = 42)
  ) +
  geom_line(
    data    = seasonal_summary,
    mapping = aes(x = season, y = avg_prcp, group = 1),
    color   = "#6B8E23", linewidth = 1.2
  ) +
  geom_point(
    data    = seasonal_summary,
    mapping = aes(x = season, y = avg_prcp, group = 1),
    color   = "#6B8E23", size = 4
  ) +
  labs(
    title = "Average Total Seasonal Precipitation",
    x     = "Season",
    y     = "Precipitation (in)"
  ) +
  seasonal_theme

# stack panels vertically using patchwork
combined_plot <- (plot_abundance / plot_precip) +
  plot_annotation(
    title   = "Seasonal Patterns in Waterfowl Abundance and Precipitation",
    caption = "Points show individual water-year seasonal totals; lines show averages across water years 2020-2025.",
    theme   = theme(
      plot.background = element_rect(fill = "#1C2B2D", color = NA),
      plot.title      = element_text(color = "#F0EAD6", size = 18, face = "bold",
                                     hjust = 0, margin = margin(b = 10)),
      plot.caption    = element_text(color = "#7A9A95", size = 8.5, hjust = 0,
                                     margin = margin(t = 8))
    )
  )

combined_plot
```